---
title: "EcoSpatial Workshop: Working with Weather Data"
author: "Sarah Goslee"
date: "2024-10-10"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    fig_width: 9
    fig_height: 7
---

The BeeShiny tool offers downloads of weather data that match the extent of the other spatial data requested.

Currently, BeeShiny only offers [monthly PRISM data](https://prism.oregonstate.edu/) on a 4km grid for mininum and maximum daily temperature and daily precipitation, but this will be expanded to daily data, and likely to other datasets and variables.

This tutorial will demonstrate:

  - loading the downloaded data into R
  - visualing the weather data using `ggplot2`
  - calculating biologically relevant indices, in this case BIOCLIM


# Packages 

This tutorial uses the dismo, SPEI, and ggplot2 packages, and requires sp and raster as dependencies.

```{r packages}

    library(ggplot2)
    library(SPEI)

```


# Data

We will compare weather for six cities across the US. The CSV file _cities.weather.csv_ can be downloaded from BeeShiny, or can be obtained from the _data_ folder in the GitHub repo.

## BeeShiny

The file _cities.csv_ should contain:

```
City,Lat,Long
Orlando,28.5384,-81.3789
StateCollege,40.7934,-77.8600
Burlington,44.4759,-73.2121
Lincoln,40.8137,-96.7026
Albuquerque,35.0844,-106.6504
Seattle,47.6061,-122.3328
```

To create the weather dataset:

  1. Access the [BeeShiny server](https://beesuite.psu.edu/beeshiny).
  2. Choose "Upload coordinates csv" and upload _cities.csv_. 
  3. Choose Long as the X coordinate column, Lat as the Y coordinate column and City as the Coordinate ID.
  4. On the Choose variables tab, select Precipitation, Maximum Temperature, Minimum temperature, and all years in the dropdown.
  5. On the Query & View Data tab, Make Query and Download selected data.
  6. Move the downloaded _data.zip_ file to your working directory. Uncompress the _data.zip_ file, and rename _weather.csv_ to _cities.weather.csv_.

If you've done step 6 or downloaded the extracted file from the GitHub repo `data` folder, use `read.csv()` to load the file directly.

```{r data1}

    weather <- read.csv("data/cities.weather.csv")

```

Bonus trick: R can read a file directly from a zipped archive, and it can also let you specify a file location. You don't need to try both (but of course you can if you want!); `weather` will be identical either way. 


```{r readzip}

    # weather <- read.csv(unz(file.choose(), "weather.csv"))

```  

## Data organization

 
The data file contains monthly temperature and precipitation for six cities, from 2008 to 2023. The plain data frame isn't the most useful way to work with time series data.

For daily data, `as.Date()` is an effective way to create a time series. Other more complex formats like _POSIXct_ are useful when time information is also needed. Since these data are monthly, let's arbitrarily choose the 15th of the month as the day of record.

Another option would be to create a variable for consecutive months from 2008-01 to 2023-12, but using a date or date/time type makes available more tools for formatting and date/time arithmetic.

By default, R plots will put the cities into alphabetical order. Turning `City` into an ordered factor will allow for custom organization.


```{r dataorganization}

    weather$date <- with(weather, as.Date(paste(year, month, "15"), format = "%Y %m %d"))


    # examples of date formatting
    # see ?strptime for a thorough list of possibilities
    head(format(weather$date, "%b %Y"))
    head(format(weather$date, "%B %Y"))

    # set order for cities
    weather$City <- factor(weather$City, ordered = TRUE, levels = with(weather, unique(City[order(Long)])))


    # daily mean temperature can be useful
    weather$tmean <- with(weather, (tmin + tmax) / 2)

```

Temperature is often displayed as a line graph, and precipitation as a bar graph showing the total. 

```{r overviewplots}

    ggplot(weather, aes(x = date, y = tmean, color = City)) +
        labs(x = "Date", y = "Temperature (C)", title = "Mean temperature") +
        geom_line()

    ggplot(weather, aes(x = date, y = pr, fill = City)) +
        labs(x = "Date", y = "Precipitation (mm)", title = "Monthly precipitation") +
        geom_bar(pos = "dodge", stat = "identity")

```

That isn't particularly helpful for comparing cities, though. 

What about a boxplot of precipitation by year?

First, calculate the annual sum of precipitation for each city and year, then plot the values.

```{r precip}

    prcp <- aggregate(pr ~ year + City, data = weather, FUN = "sum")

    ggplot(prcp, aes(x = City, y = pr, fill = City)) +
        labs(x = "Date", y = "Precipitation (mm)", title = "Annual precipitation") +
        geom_boxplot()

```

A third common display for weather data is a combined plot of mean temperature and precipitation together. That takes a few more steps in `ggplot2`, but is entirely feasible. The trickiest part putting both on the same scale to get the double y axes.

```{r combined}

    monthly <- aggregate(. ~ month + City, data = subset(weather, select = c(City, month, pr, tmax, tmin)), FUN = "mean")

    temp.range <- with(monthly, range(c(tmin, tmax)))
    pr.range   <- range(monthly$pr)

    b <- diff(pr.range)/diff(temp.range)
    a <- pr.range[1] - b * temp.range[1]

    ggplot(monthly) +
        geom_bar(aes(x =  month, y = pr), stat = "identity", fill = "darkgreen") +
        geom_line(aes(x = month, y = a + tmin * b, color = "blue", linewidth = 2)) +
        geom_line(aes(x = month, y = a + tmax * b, color = "red", linewidth = 2)) +
        scale_y_continuous("Precipitation (mm)", sec.axis = sec_axis(~ (. - a)/b, name = "Temperature (C)")) +
        scale_x_continuous("Month", breaks = 1:12, labels = month.abb) +
        facet_wrap( ~City) +
        theme(legend.position = "none")

```


# Weather and climate indices

Often simply temperature and precipitation are not useful enough, and we want some indices that are biologically relevant for the organisms of interest. 

## BIOCLIM

One commonly-used set of indices is the BIOCLIM variables, which are:


  - bc01	Annual Mean Temperature
  - bc02	Mean Diurnal Range (Mean of monthly (max temp - min temp))
  - bc03	Isothermality (BIO2/BIO7) (* 100)
  - bc04	Temperature Seasonality (standard deviation * 100)
  - bc05	Max Temperature of Warmest Month
  - bc06	Min Temperature of Coldest Month
  - bc07	Temperature Annual Range (BIO5-BIO6)
  - bc08	Mean Temperature of Wettest Quarter
  - bc09	Mean Temperature of Driest Quarter
  - bc10	Mean Temperature of Warmest Quarter
  - bc11	Mean Temperature of Coldest Quarter
  - bc12	Annual Precipitation
  - bc13	Precipitation of Wettest Month
  - bc14	Precipitation of Driest Month
  - bc15	Precipitation Seasonality (Coefficient of Variation)
  - bc16	Precipitation of Wettest Quarter
  - bc17	Precipitation of Driest Quarter
  - bc18	Precipitation of Warmest Quarter
  - bc19	Precipitation of Coldest Quarter

where a "quarter" is any consecutive three months. Most commonly, each year is wrapped around, so a quarter might be Dec-Jan-Feb, where December and January are both 2023, rather than December 2023 and January 2024. That makes it possible to calculate these indices for a single year. It's what we will do here, as is conventional, but you might want to use continuous months for other applications.

I previously used the [climates](https://github.com/jjvanderwal/climates) package, but it has as dependencies several packages that are no longer on CRAN, so I've provided you a function.

There's another package function, `dismo::bioclim()` that does something related but different.


```{r bioclim}

    source("bioclim.R")

    # calculate bioclim variables for one city
    statecollege.bioclim <- bioclim(subset(weather, City == "StateCollege"))

    ###

    # calculate bioclim variables for each city
    all.bioclim <- lapply(split(weather, weather$City), bioclim)

    # create a long-form data frame for plotting
    all.bioclim <- lapply(seq_along(all.bioclim), function(i) {
        all.bioclim[[i]]$City <- names(all.bioclim)[i]
        all.bioclim[[i]]
    })
    all.bioclim <- do.call(rbind.data.frame, all.bioclim)
    all.bioclim$City <- factor(all.bioclim$City, ordered = TRUE, levels = levels(weather$City))

    all.bioclim.long <- tidyr::pivot_longer(all.bioclim, cols = 2:20)
    

    # plot the variables for each city
    ggplot(all.bioclim.long, aes(x = City, y = value, fill = City)) +
        labs(x = "City", y = "BIOCLIM Value") +
        geom_boxplot() + 
        facet_wrap( ~name, scales = "free") +
        theme(axis.text.x = element_blank())


```


## SPEI

Drought is a concern in many parts of the US (alongside flooding!), and the [SPEI drought index](https://spei.csic.es/) is commonly used to quantify drought at multiple timescales, here 1, 6, and 12 months lag.

The SPEI package includes functions for calculating SPEI, as well as several potential evapotranspiration methods and a few other things.

```{r spei, results='hide'}
     
    # for each city, calculate PET and water balance
    # and then three different SPEI values
    # save the SPEI values in a list
    
    weather$PET <- NA
    weather$BAL <- NA
    weather$SPEI01 <- weather$SPEI06 <- weather$SPEI12 <- NA

    for(thiscity in unique(weather$City)) {

        # simplest form of PET calculation
        thispet <- with(subset(weather, City == thiscity), 
                thornthwaite(tmean, Lat[1]))
        thisbal <- with(subset(weather, City == thiscity),
                pr - thispet)

        # save those results
        weather$PET[weather$City == thiscity] <- thispet
        weather$BAL[weather$City == thiscity] <- thisbal

        # calculate the SPEI values
        thisspei01 <- spei(thisbal,  1)
        thisspei03 <- spei(thisbal,  3)
        thisspei12 <- spei(thisbal, 12)
        thisspei36 <- spei(thisbal, 36)

        # save just the fitted values for plotting
        weather$SPEI01[weather$City == thiscity] <- thisspei01$fitted
        weather$SPEI03[weather$City == thiscity] <- thisspei03$fitted
        weather$SPEI12[weather$City == thiscity] <- thisspei12$fitted

    }

    # clean up, if desired
    rm(list = ls(pattern = "^this"))

```


The plot method for SPEI uses base graphics rather than `ggplot2`. 

*Note:* The base plot method is currently broken, so the above code chunk saves the fitted values from the SPEI objects so we can use `ggplot2` for graphics. 


```{r speiplot}

    # add a column for positive/negative SPEI for coloring the bar plots
    weather$POS01 <- weather$SPEI01 >= 0
    weather$POS03 <- weather$SPEI03 >= 0
    weather$POS12 <- weather$SPEI12 >= 0



    ggplot(weather, aes(x = date, y = SPEI01, fill = POS01)) +
        labs(x = "Date", y = "SPEI", title = "SPEI: 1 Month Lag") +
        geom_bar(pos = "dodge", stat = "identity") + 
        facet_wrap( ~City) +
        theme(legend.position = "none")

    ggplot(weather, aes(x = date, y = SPEI12, fill = POS12)) +
        labs(x = "Date", y = "SPEI", title = "SPEI: 12 Month Lag") +
        geom_bar(pos = "dodge", stat = "identity") + 
        facet_wrap( ~City) +
        theme(legend.position = "none")
    
```

# The importance of scale

But... I thought the Southwest was experiencing a prolonged drought?

Yes, but from the perspective of the 2008-2023 timeperiod, not as noticeably. 

A longer timespan might show something different. This example also demonstrates how to aggregate daily to monthly values before calculating SPEI or BIOCLIM indices.

The file _abq.ghcn.RDS_ contains daily weather data for the Albuquerque, NM airport from [GHCN](https://www.ncei.noaa.gov/products/land-based-station/global-historical-climatology-network-daily). The column headings are the GNCN standard, but extraneous columns have been removed and the data trimmed to 1932-2023.


```{r abq, results='hide'}

    # load and clean up data
    abq <- readRDS("abq.ghcn.RDS")
    abq$TMEAN.VALUE <- with(abq, (TMIN.VALUE + TMAX.VALUE) / 2)

    # precipitation is sum; temperature is mean!
    abq.monthly.prcp <- aggregate(PRCP.VALUE ~ MONTH + YEAR, data = abq, FUN = "sum", na.rm = TRUE)
    abq.monthly.temp <- aggregate(TMEAN.VALUE ~ MONTH + YEAR, data = abq, FUN = "mean", na.rm = TRUE)

    thispet <- thornthwaite(abq.monthly.temp$TMEAN.VALUE, lat = 35.0419)
    thisbal <- abq.monthly.prcp$PRCP.VALUE - thispet

    abq.spei12 <- spei(thisbal, 12)


    abq.monthly <- data.frame(abq.monthly.temp, PRCP.VALUE = abq.monthly.prcp$PRCP.VALUE,
        SPEI12 = abq.spei12$fitted,
        POS12 = abq.spei12$fitted >= 0)
    abq.monthly$date <- with(abq.monthly, as.Date(paste(YEAR, MONTH, "15"), format = "%Y %m %d"))

    ggplot(abq.monthly, aes(x = date, y = SPEI12, fill = POS12)) +
        labs(x = "Date", y = "SPEI", title = "Albuquerque SPEI: 12 Month Lag") +
        geom_bar(pos = "dodge", stat = "identity") + 
        theme(legend.position = "none")


```


# Check your learning

  - Calculate monthly minimum and maximum temperatures for Albuquerque.
  - Plot the monthly weather variables for Albuquerque.
  - Calculate the BIOCLIM variables for the longer Albuquerque time series. There is no help file, but you can use 
    `args(bioclim)` to see the function arguments. Pay attention to the column names!

    
