---
title: "EcoSpatial Workshop"
author: "Heather Grab, Kevin Li, Sarah Goslee"
date: "2024-10-03"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---


## Workshop Outline {-}
This workshop is designed to introduce users to the new beeshiny R shiny application. 

beeshiny can be used to query data commonly used when modeling the environemental drivers of species' or commmunity responses. beeshiny was originally build as a companion to [Beescape](https://beescape.psu.edu/). 

beeshiny allows users to extract:

* climate data: monthly precipitation, min, and max temperature from PRISM
* crop land covers: from the NASS CropLand Data Layer
* beescape indices: spring forage, summer forage, fall forage, nesting, and pesticides

### Load packages
There are a number of packages in R that are helpful in working with spatial data. For this workshop we will be using [sf](https://r-spatial.github.io/sf/) and [terra](https://rspatial.org/index.html). 
For data wrangling and visualizations we will use `dplyr` and `ggplot2` respectively. 

Because some spatial functions have the same names as dplyr functions it is helpful to load the spatial packages last. We can also use the `::` to specify the package for function calls. 

Unhash and use install.packages for any packages your may not already have installed
```{r packages, echo=TRUE, results='hide', message=FALSE}
library(dplyr)
library(ggplot2)
#install.packages("sf", dependencies = TRUE)
#install.packages("terra", dependencies = TRUE)
library(sf)
library(terra)
```


## Get raster data from beeshiny
Use beeshiny to download a CDL raster for any county in the US you would like for any year you want

## Working with raster data in R

Downloaded data from beeshiny is packaged as a zipped file called data.zip. Inside this file yYou should now have a .tif file. The name indicates the data type, in this case CDL, the year and the FIPS code that corresponds to the county you selected. 

Move the .tif file into the /data folder in your ecospatial-workshop directory

The first thing we will do is read 


```{r}

```

### Check your learning

Use beeshiny to download the CDL raster for Centre County Pa for any year you want

Move the CDL raster for Centre County to the /data folder in your ecospatial-workshop directory

Read the raster into R as a terra

```{r}

```



## Working with point data



```{r}

```


## Working with climate data



```{r}

```

