---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE
)
```

# EcoSpatial Summit Workshop

In preparation for the EcoSpatial Summit Workshop, please download the following programs:

* [R](https://cran.rstudio.com/)

* [RStudio](https://posit.co/download/rstudio-desktop/)

## Load packages

There are a number of packages in R that are helpful in working with spatial data. For this workshop we will be using `sf` and `terra`. 
For data wrangling and visualizations we will use `dplyr`, `ggplot2` and `tidyterra`. 

| Name | Description | Link |
|:--|:--|:--|
| `dplyr` | Package that provides a 'grammar' of data manipulation in `R` | https://dplyr.tidyverse.org/ |
| `ggplot2` | Package that provides a system for declaratively creating graphics | https://ggplot2.tidyverse.org/ |
| `sf` | Package for manipulating 2-D geographic vector data | https://r-spatial.github.io/sf/ |
| `terra` | Package for spatial data analysis | https://rspatial.github.io/terra/ |
| `tidyterra` | Package for integrating objects from `terra` with the `dplyr` and `ggplot2` packages | https://dieghernan.github.io/tidyterra/ |

Unhash (remove the preceding `#`) and use install.packages for any packages your may not already have installed

```{r packages, echo=TRUE, results='hide', message=FALSE}
# install.packages("dplyr", dependencies = TRUE)
# install.packages("ggplot2", dependencies = TRUE)
# install.packages("sf", dependencies = TRUE)
# install.packages("terra", dependencies = TRUE)
# install.packages("tidyterra", dependencies = TRUE)

library(dplyr)
library(ggplot2)
library(sf)
library(terra)
library(tidyterra)
```
