---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE
)
```

# EcoSpatial Summit Workshop

This workshop is designed to introduce users to the new [**beeshiny app**](https://beesuite.psu.edu/beeshiny/). 

beeshiny can be used to query data commonly used when modeling the environemental drivers of species' or commmunity responses. beeshiny was originally build as a companion to [Beescape](https://beescape.psu.edu/). 

beeshiny allows users to extract:

* climate data: monthly precipitation, min, and max temperature from PRISM
* crop land covers: from the NASS CropLand Data Layer
* beescape indices: spring forage, summer forage, fall forage, nesting, and pesticides

## Nomenclature

* **raster:** A form of geographic data that represents the landscape as a grid of cells. Grid values can be continuous values, signifying things like elevation, temperature, or spectral information (e.g. color) in a satellite image; or categorical values, signifying thematic classes like land cover.
* **vector:** Another form of geographic data that represents specific features on the landscape composed of discrete geometric locations that consist of x and y values. These can be points, lines, or polygons.
* **Cropland Data Layer (CDL):** A set of raster datasets produced by the USDA Nationa Agricultural Statistics (NASS) Service and Agricultural Research Service (ARS) that represents annual crop-specific land cover produced from satellite imagery and ground reference data.

## Data Details

* We will use data downloaded from the [**beeshiny app**](https://beesuite.psu.edu/beeshiny/).

* We will also use point data representing hypothetical sites in Centre County, PA, which can be downloaded from this repository at `/data/centre_co_pts.csv`.

 * *Note that if you are unable to access the data on beeshiny, backup data are available in the repository in* `data/backups`

## Load packages

There are a number of packages in R that are helpful in working with spatial data. For this workshop we will be using `sf` and `terra`. 
For data wrangling and visualizations we will use `dplyr`, `ggplot2` and `tidyterra`. 

| Name | Description | Link |
|:--|:--|:--|
| `dplyr` | Package that provides a 'grammar' of data manipulation in `R` | https://dplyr.tidyverse.org/ |
| `ggplot2` | Package that provides a system for declaratively creating graphics | https://ggplot2.tidyverse.org/ |
| `sf` | Package for manipulating 2-D geographic vector data | https://r-spatial.github.io/sf/ |
| `terra` | Package for spatial data analysis | https://rspatial.github.io/terra/ |
| `tidyterra` | Package for integrating objects from `terra` with the `dplyr` and `ggplot2` packages | https://dieghernan.github.io/tidyterra/ |

Because some spatial functions have the same names as dplyr functions it is helpful to load the spatial packages last. We can also use the `::` to specify the package for function calls. 

Unhash (remove the preceding `#`) and use install.packages for any packages your may not already have installed

```{r packages, echo=TRUE, results='hide', message=FALSE}
# install.packages("dplyr", dependencies = TRUE)
# install.packages("ggplot2", dependencies = TRUE)
# install.packages("sf", dependencies = TRUE)
# install.packages("terra", dependencies = TRUE)
# install.packages("tidyterra", dependencies = TRUE)

library(dplyr)
library(ggplot2)
library(sf)
library(terra)
library(tidyterra)
```

# Working with raster data in R

## Get raster data from beeshiny

Use [**beeshiny**](https://beesuite.psu.edu/beeshiny/) to download the cropland data layer (CDL) raster for **Centre County, Pennsylvania** for the year **2021**.

## Read in raster data

Downloaded data from beeshiny is packaged as a zipped file called data.zip. Inside this file you should have a .tif file. The name indicates the data type, in this case CDL, the year and the FIPS code that corresponds to the county you selected. 

**For the code below to work, you must move the .tif file into the `/data` folder in your ecospatial-workshop directory.**

The first thing we will do is use the `rast()` function to read the .tif file into R as a `SpatRaster` object. NOTE: if you aren't able to download the raster from beeshiny, you can access a pre-downloaded copy in the `data/backups/` folder.

```{r create SpatRaster of cdl}
centre_cdl <- rast("data/CDL_2021_FIPS_42027.tif")    # add `backup/` after data to use backup data

centre_cdl
```
 
An important attribute of spatial data are their Coordinate Reference System or CRS. This information tells us what model of the earth (ex WGS84 or NAD83) is being referenced as well as the units of the coordinates such as decimal degrees. 

Rasters downloaded from beeshiny inherit their CRS from the raster they were originally extracted from.

Let's view the CRS for your county cdl:

```{r crs}
crs(centre_cdl)
```

We can see that our CDL raster is using the North American Datum of 1983 as its model for the shape of the earth. Our two-dimensional projection model of earth's 3d surface is Alber's Equal Area and the units of our coordinates are in meters.

## Visualize raster data

We can visualize our county CDl raster by using the base R `plot` function:

```{r base}
plot(centre_cdl)
```

We'll talk about other plotting options such as using the packages `ggplot2` and `tidyterra` a little later.

### Reclass to named CDL classes

Right now we are viewing the raw numeric codes of the CDL. These codes mean something: the crop land cover class. We can tell `terra` what these codes mean so that it will show the land cover class names rather than the raw values. The information for these land cover values (along with their colors) are stored in the color table in `data/cdl_colormap.csv`.

```{r load_coltab}
cdl_colormap <- read.csv("data/cdl_colormap.csv")   # read in the table

head(cdl_colormap)  # use `head()` to take a look at the first 5 rows of cdl_colormap
```

**Question:** Which column represents the raw CDL values and which column represents their land cover class?

We set the levels of the raster to the land cover class names using the relevant elements of `cdl_colormap`: 'value' (column 1) and 'class_name' (column 6).

```{r def_level}
levels(centre_cdl) <- cdl_colormap[,c(1,6)]

plot(centre_cdl)  # plot, as above, but now R knows what the numeric values mean
```

Then we can recolor the classes to match the traditional [NASS CDL style](https://www.nass.usda.gov/Research_and_Science/Cropland/docs/US_2023_CDL_legend.jpg).

This color information is stored in columns 2-5 of the `cdl_colormap` table, representing red, green, blue, and alpha (transparency) values. We use the function `coltab` to supply the color map with this information in columns that are in this specific order, plus the corresponding (raw) raster value as the first column (they are already set up as the first 5 columns of `cdl_colormap`)

```{r }
coltab(centre_cdl) <- cdl_colormap[,1:5]

plot(centre_cdl)  # plot, as above, but now R knows what the numeric values mean and assigns colors
```

Let's look at some customization options. To do this, we'll be using the `ggplot2` `tidyterra` packages. `ggplot2` plots a little differently from base R. Data elements are layered on top of a base plot using `+` signs and functions that start with `geom_`. Here, we're starting with a blank base plot and adding the spatraster object using `geom_spatraster()`.

```{r ggplot}
ggplot() +  # blank base plot
  geom_spatraster(data = centre_cdl, aes(fill = class_name))   # add the spatraster layer
```

The legend is now quite large but can be easily modified using `ggplot2` `theme` functions.

```{r}
ggplot() +
  geom_spatraster(data = centre_cdl, aes(fill = class_name)) +
  theme(legend.title = element_text(size = 7), # make legend title smaller
        legend.text = element_text(size = 7), # make legend text smaller
        legend.key.size = unit(0.25, 'cm'), #make legend color keys smaller
        legend.position="bottom") # move legend to the bottom of the plot
```

### Reclass crop land cover to spring floral resources

Up to know, our CDL raster values are still numeric values that represent crop land cover class from the CDL. We can reclassify these CDL values to the estimated floral resources of each land cover class, based on [Koh et al. (2015)](https://www.pnas.org/doi/abs/10.1073/pnas.1517685113).

A reclassification table based on Koh et al. is in the `data/` folder. The table rows connect each CDL value to its corresponding class name and the values for several indices. 

```{r read in koh table}
reclass_table <- read.csv("data/cdl_reclass_koh.csv")  # read in the reclassification table

head(reclass_table)     # take a look at the first 5 rows
```

A reclassification table assigns the original values of a raster (listed in the first column) to a new value (listed in the second column). This is done using the `classify()` function. 

We will select the columns corresponding to the CDL value and the spring floral resources as our original and new values, respectively, for our reclass table. We'll reclassify the Centre County CDL and generate a map of spring floral resources across the county.

```{r reclass to forage}
centre_floral_sp <- classify(centre_cdl,
                    reclass_table[,c("value",
                                     "floral_resources_spring_index")])

plot(centre_floral_sp)
```

## Inspect raster values

Using the `values()` function we can directly inspect the spring floral values for Centre County. We will set the argument `na.rm=TRUE` so that all the empty cells (outside of the county) are not included. The result of `values()` shows individual grid cell values. In this case we will only extract the first 20 grid cell values.

```{r inspect_values}
values(centre_floral_sp, na.rm=TRUE)[1:20]   # just the first 20 cells
```

We can also use some basic summary functions to view the distribution of floral resource values for the county.

```{r summarize_values}
summary(values(centre_floral_sp, na.rm=TRUE))    # make a summary with the quartiles and the mean

hist(values(centre_floral_sp, na.rm=TRUE))       # make a basic histogram of values
```

## Write out raster files

We can save our raster files as a .tif using `writeRaster`. For the CDL raster we will save the data in "INT1U" format which will also save the class names and color table. 

```{r write_raster}
writeRaster(centre_cdl, "data/centre_county_cdl_2015.tif", overwrite=TRUE, datatype="INT1U")

writeRaster(centre_floral_sp, "data/centre_county_springfloral_2015.tif", overwrite=TRUE)
```

## Check your learning

1. Use beeshiny to download the CDL raster for another county in the continental US for any year you want.

2. Move the CDL raster for Centre County to the `/data` folder in your ecospatial-workshop directory.

3. Read the raster into R as a terra `SpatRaster` using `rast()`.

4. Assign the CDL value codes with land cover class names using `levels()` and assign colors using `cdl_colormap()`.

```{r check learning raster, class.source = 'fold-hide', message=FALSE}

# county_cdl <- rast("data/CDL_2021_FIPS_42027.tif") 

# levels(county_cdl) <- cdl_colormap[,c(1,6)]

# coltab(county_cdl) <- cdl_colormap[,1:5]

# plot(county_cdl)

```

# Working with vector (points and polygons) data

## Read in point data

Often we collect and store point data as coordinates in a non-spatial format such as a csv file which we need to convert into a spatial object in R. 

In the /data directory you should find a csv file with a set of points from Centre County Pa. We will read in and inspect these data.
```{r read in pts}
centre_sites <- read.csv("data/centre_co_pts.csv")
centre_sites
```

and convert them into an sf spatial object. We use the `coords=` argument to indicate the columns that hold the longitude (X) and latitude (Y). We can also specify the CRS using the 'crs=' arguments and the appropriate EPSG code or other crs object. In our case we will give the EPSG code 4326 which corresponds to WGS 84, the system commonly used by Google Earth. 
```{r make pts spatial}
centre_sites <- st_as_sf(centre_sites, 
                         coords=c("Long","Lat"), #indicate the x and y columns
                         crs = 4326) #set the crs
```

when we inspect the center_sites object we see that there is now additional information at the top. 
```{r inspect sf pts}
centre_sites
```


## Visualize point data

Using `ggplot2` and `sf` we can add our points to our earlier map of Centre county with the `geom_sf()` function. 

```{r}
#Centre county CDL from raster exercise
centre_cdl <- rast("data/centre_county_cdl_2015.tif") 

ggplot()+
  geom_spatraster(data = centre_cdl, aes(fill = class_name)) +
  geom_sf(data=centre_sites) + 
  theme(legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7), 
        legend.key.size = unit(0.25, 'cm'),
        legend.position="bottom") 
```

## Buffer around points

We can create a polygon object from our points by buffering around them in a 1000m radius. For example, we might do this when assessing land cover values around a site.
```{r}
centre_sites_1000m <- st_buffer(centre_sites, 1000)
centre_sites_1000m
```


We can visualize this buffer by adding it to our plot. Note you may need to expand your plot size to see the buffers behind the points. 
```{r}
ggplot()+
  geom_spatraster(data = centre_cdl, aes(fill = class_name)) +
  geom_sf(data = centre_sites_1000m)+
  geom_sf(data=centre_sites) + 
  theme(legend.title = element_text(size = 7), 
        legend.text = element_text(size = 7), 
        legend.key.size = unit(0.25, 'cm'),
        legend.position="bottom") 
```


## Extract CDL values within buffers

If we wanted to assess land cover around a site in a way that is similar to the results that beeshiny provides, we can use the `extract()` function from terra. First providing the raster we want to extract values from, then the sf object that defines the zones in which the raster will the summarized. 
```{r extract}
centre_table_1000m <- extract(centre_cdl, centre_sites_1000m, fun="table", ID=F)
centre_table_1000m
```

We can calculate additional statistics such as the area within the buffer for each site. 
```{r area}
Area_m2 <- rowSums(centre_table_1000m)*900
```

and turn our table into proportional cover of different land classes. 
```{r}
centre_table_1000m_prop <- centre_table_1000m/rowSums(centre_table_1000m)

centre_table_1000m_prop <- cbind(Area_m2,centre_table_1000m_prop)
centre_table_1000m_prop
```

## Write out shapefiles
we can write out vector data to shapefiles using `st_write()`
```{r}
st_write(centre_sites, "data/centre_sites.shp")
st_write(centre_sites_1000m, "data/centre_sites_buffer_1000m.shp")
```



## Check your learning

1. Using beeshiny, extract the CDL values for 1km around the centre county points for the same year of the CDL that you used in the last activity. 

2. Compare the CDL values from beeshiny for the centre county sites to the values you produced using the extract function above. Values should be similar but not exacly the same because beeshiny uses exactextractr (`?exact_extract`) instead of the `terra::extract` function. 

# Working with climate data

```{r}

```

# Bringing it all together
```{r}

```

